Baseline centering problem

### 2019-01-14 paper/tangotd.tex

Building a universal policy. One may make the policy depend on a goal
representation $\omega$ as an additional input. One may simultaneously
compute $Z(s_1,s_2;\omega)$, namely the $Z$ function when following the
policy with goal $\omega$, and train at the same time a map $s_2\mapsto \omega=\Omega(s_2)$ and
policy parameters $\pi(a|s_1,\omega)$ such that following
$\pi(|\Omega(s_2))$ indeed goes towards $s_2$, namely, the policy
gradient for every $s_2$ is
\begin{equation}
\partial \pi(a|s,\Omega(s_2))\E_{s'|s,a} \E_{s_2\sim \rho} \,
(z(s',s_2;\Omega(s_2))-z(s,s_2;\Omega(s_2)))
\end{equation}
This equation is used to train the policy parameters; $\Omega(s_2)$ is
trained so that $\omega=\Omega(s_2)$ indeed maximizes $z(s,s_2;\omega)$
(because the best way to reach $s_2$ is to select the $\omega$ with
maximal $z(s,s_2;\omega)$ at each step; TODO: from which $s$? TODO: here
I'm confusing constant action plans and time-variable action plans).
[TODO check this is indeed the policy gradient for the time spent at
$s_2$ if following policy $\pi(\Omega(s_2))$]. Advantage: diversity of
policies should appear by itself, no need to manually force options to be
distinct. Depends on a chosen distribution of goals $s_2$; we could learn
to go to uniformly distributed states, or create a goal at points where
we observe a high reward value (so learning to go to all states with a
reward). Just fixing $\Omega(s)\deq s$ once and for all might work, in
principle, but may be ambitious (or use a summary/auto-encoded value of
$s$). TODO check:
this is with a reward a Dirac at $s_2$, check there's no technical
problem wrt the reference distribution $\rho(s_2)$, as we use $z$ not $Z$
in the policy gradient and $z$ is the density of $Z$ wrt $\rho$ but I
think it's correct. Actually, for goal representations, a goal is not
necessarily a state itself, should be a feature of states, or some
representation of a reward function, so we should figure out the way to
give a meaning to $\Omega(R)$ for an arbitrary reward $R$. The
representation of $Z$ as $Z=\transp{\green}f$ does that automatically,
thanks to linearity wrt $R$. This could even help with exploration:
setting temporary goals towards already-seen past states solves the "pure
random walk" problem by introducing temporary momentum; one can even
increase temperature a bit on $\omega$, by using slightly larger/more
variant values of $\omega$ as the goal.

@@

What I don't like here: This still does not capture the composition of
policies/goals (like: to go to $s_2$, first go to $s_1$ via the
$s_1$-policy, then apply the known $s_2$-policy from $s_1$). The
Newton--Bellman equation captures, \emph{for a fixed policy}, the
composition of paths of that policy. But the above still treats
$z(s_1,s_2;\omega)$ as a separate problem for each $\omega$.

@@

By multiplying \eqref{eq:optleftbellman} by $V$ on the left, one finds the
Newton--Bellman equation for the optimal function $V$ seen as a matrix:
\begin{equation}
V^2_{s_1s_2}=V_{s_1s_2}+\gamma \sum_{s_3} V_{s_1s_3}\max_a
\E_{s_4|s_3,a} V_{s_4s_2}
\end{equation}
and representing $V$ by its density wrt $\rho$ namely
$V_{s_1s_2}=z^\ast(s_1,s_2)\rho(s_2)$ this suggests the Newton method
\begin{multline}
z^\ast(s_1,s_2)\gets 2z^\ast(s_1,s_2)-\E_{s_3\sim \rho}\,
z^\ast(s_1,s_3)z^\ast(s_3,s_2)\\
+\gamma
\E_{s_3\sim \rho} \,z^\ast(s_1,s_3)\max_a \E_{s_4\sim P(s_4|s_3,a)} z^\ast(s_4,s_2)
\end{multline}
satisfied by the optimal $z$ function.

>>>>

Building a universal policy. One may make the policy depend on a goal
representation $\omega$ as an additional input. One may simultaneously
compute $Z(s_1,s_2;\omega)$, namely the $Z$ function when following the
policy with goal $\omega$, and train at the same time a map $s_2\mapsto \omega=\Omega(s_2)$ and
policy parameters $\pi(a|s_1,\omega)$ such that following
$\pi(|\Omega(s_2))$ indeed goes towards $s_2$, namely, the policy
gradient for every $s_2$ is
\begin{equation}
\partial \pi(a|s,\Omega(s_2))\E_{s'|s,a} \E_{s_2\sim \rho} \,
(z(s',s_2;\Omega(s_2))-z(s,s_2;\Omega(s_2)))
\end{equation}
This equation is used to train the policy parameters; $\Omega(s_2)$ is
trained so that $\omega=\Omega(s_2)$ indeed maximizes $z(s,s_2;\omega)$
(because the best way to reach $s_2$ is to select the $\omega$ with
maximal $z(s,s_2;\omega)$ at each step; TODO: from which $s$? TODO: here
I'm confusing constant action plans and time-variable action plans).
[TODO check this is indeed the policy gradient for the time spent at
$s_2$ if following policy $\pi(\Omega(s_2))$]. Advantage: diversity of
policies should appear by itself, no need to manually force options to be
distinct. This could even help with exploration:
setting temporary goals towards already-seen past states solves the "pure
random walk" problem by introducing temporary momentum; one can even
increase temperature a bit on $\omega$, by using slightly larger/more
variant values of $\omega$ as the goal. Depends on a chosen distribution of goals $s_2$; we could learn
to go to uniformly distributed states, or create a goal at points where
we observe a high reward value (so learning to go to all states with a
reward). Just fixing $\Omega(s)\deq s$ once and for all might work, in
principle, but may be ambitious (or use a summary/auto-encoded value of
$s$). TODO check:
this is with a reward a Dirac at $s_2$, check there's no technical
problem wrt the reference distribution $\rho(s_2)$, as we use $z$ not $Z$
in the policy gradient and $z$ is the density of $Z$ wrt $\rho$ but I
think it's correct. Actually, for goal representations, a goal is not
necessarily a state itself, should be a feature of states, or some
representation of a reward function, so we should figure out the way to
give a meaning to $\Omega(R)$ for an arbitrary reward $R$. The
representation of $Z$ as $Z=\transp{\green}f$ does that automatically,
thanks to linearity wrt $R$.

@@

What I don't like here: This still does not capture the composition of
policies/goals (like: to go to $s_2$, first go to $s_1$ via the
$s_1$-policy, then apply the known $s_2$-policy from $s_1$). The
Newton--Bellman equation captures, \emph{for a fixed policy}, the
composition of paths of that policy. But the above still treats
$z(s_1,s_2;\omega)$ as a separate problem for each $\omega$. The optimal
Newton--Bellman equation for $z^\ast$ (see below) does capture
composition of optimal actions, though.

@@

By multiplying \eqref{eq:optleftbellman} by $V$ on the left, one finds the
Newton--Bellman equation for the optimal function $V$ seen as a matrix:
\begin{equation}
V^2_{s_1s_2}=V_{s_1s_2}+\gamma \sum_{s_3} V_{s_1s_3}\max_a
\E_{s_4|s_3,a} V_{s_4s_2}
\end{equation}
and representing $V$ by its density wrt $\rho$ namely
$V_{s_1s_2}=z^\ast(s_1,s_2)\rho(s_2)$ this suggests the Newton method
\begin{multline}
z^\ast(s_1,s_2)\gets 2z^\ast(s_1,s_2)-\E_{s_3\sim \rho}\,
z^\ast(s_1,s_3)z^\ast(s_3,s_2)\\
+\gamma
\E_{s_3\sim \rho} \,z^\ast(s_1,s_3)\max_a \E_{s_4|s_3,a} z^\ast(s_4,s_2)
\end{multline}
satisfied by the optimal $z$ function. This is not quite an instance of
the Newton method for matrix inversion anymore (because the operator
$\Id-\gamma \max_a \E_{\cdot|a}$ is not linear), so convergence
is unclear.

Moreover, this matrix $V$ does not contain all the information about
optimizing rewards: the optimal strategy to optimize a reward $R$ is not
an average of the optimal strategies for going to $s_2$ weighted by
$R(s_2)$...
